---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: svm-latex-ms.tex
title: "SWATplusR 0.2.4: Running SWAT2012 and SWAT+ Projects in R"
thanks: "This fact sheet refers to the current version 0.2.4 (https://doi.org/10.5281/zenodo.2630510)"
author:
- name: Christoph Schürz
  affiliation: BOKU University, Vienna, Austria. christoph.schuerz@boku.ac.at
abstract: "The `SWATplusR` package provides tools to link existing SWAT2012 and SWAT+ models with your modeling workflows in R. `SWATplusR` enables you to execute SWAT simulations and to control all relevant parameters of a SWAT simulation, such as changes in model parameters, the simulation periods and time steps, or the simulated variables that should be returned to R. The central goal of `SWATplusR` is to return simulation results in a *tidy* format to facilitate an easy implementation of SWAT simulations, together with other R packages into clean and efficient R programming workflows. To efficiently handle large SWAT projects with large numbers of model evaluations and/or large simulation outputs, `SWATplusR` provides parallel computation and incremental saving and selective loading of simulation results into and from SQLite data bases."
# keywords: "SWAT, R"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
# spacing: double
# bibliography: ~/Dropbox/master.bib
biblio-style: apsr
endnote: no
---

# Installation

`SWATplusR` is currently under development. You can install an unreleased version of `SWATplusR` from the package's GitHub repository.

``` {r, eval = FALSE}
# If you do not have the package devtools installed
install.packages("devtools")

# If access is denied, then the repository might be still set to private
# In that case, please contact me (c.schuerz@posteo.org).
devtools::install_github("chrisschuerz/SWATplusR")
```


# Functionality and workflow

The functionality of `SWATplusR` is reduced to a few essential functions that maintain the link between the SWAT project on the local hard drive and the R environment. With `load_demo()` you can retrieve demo data sets of SWAT projects, calibration data, and shape files of the demo catchment. With `run_swat2012()` and `run_swat2012()` you can run a SWAT model located in a local project folder and return simulation outputs to R that were defined with `define_output()`. Simulation results can be saved incrementally to an SQLite data base when a `save_file` is defined in `run_swat*()`. With `load_swat_run()` all or selected parts of the simulation results stored in the data base can be loaded back to R. `scan_swat_run()` scans the content of saved simulations and returns meta data on the saved content.



```{r, echo=FALSE, }
knitr::include_graphics(here::here("man/figures/workflow.pdf"))
```

\vspace*{25px}


# Getting started with `SWATplusR`

You can explore the basic functionality of `SWATplusR` in the [Get started](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#loading-demos) section. There you can learn the following basics:

- Loading demo data [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#figures)
    * SWAT projects [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.htmll#swat-projects)
    * Observation data [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#observation-data)
    * Spatial catchment data [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#spatial-data)


- Performing first SWAT model runs from R [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#first-swat-model-runs)
    * Output definition [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#output-definition)
    * Exploring the first simulation results [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#exploring-a-simulation)

- Defining parameter modifications for a SWAT run [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#model-parameter-alteration)
    * Simulation with a single parameter set [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#simulation-with-a-single-parameter-set)
    * Simulations with many parameter sets [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#simulations-with-many-parameter-sets)
    * Exploring the outputs when the parameter were modified [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#exploring-simulations-with-parameter-alterations)

- Saving and loading SWAT simulations [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#saving-and-loading-simulations)
- Further input arguments of `SWATplusR` [>>](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#further-swatplusr-input-arguments)

The [Articles](https://chrisschuerz.github.io/SWATplusR/articles/) section is a collection of tutorials for typical topics on SWAT modeling. Here you can learn how to use `SWATplusR` in combination with other R packages to perform tasks such as:

- Parameter sensitivity analysis [>>](https://chrisschuerz.github.io/SWATplusR/articles/par_sensitivity.html)
- Model parameter optimization [>>](https://chrisschuerz.github.io/SWATplusR/articles/par_optim.html)
- Parameter sampling and model calibration [>>](https://chrisschuerz.github.io/SWATplusR/articles/par_sampl_calib.html)
- Visualization >>

The Articles section will be updated in the future with further topics that can be relevant for any modeling workflow with `SWATplusR`.

# Reporting bugs

This is the first release of the R package. If you encounter any issues with the package, please report the problems (in the best case with a small reproducible example) on the package's [issues](https://github.com/chrisschuerz/SWATplusR/issues) page.

# Citation

Please cite the package when you use it in your work. You can find the bibtex entry for the package [here](https://chrisschuerz.github.io/SWATplusR/authors.html). 
