---
title: "02 - Visualization of simulation results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring SWATplusR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: literature.bib
link-citations: yes
csl: copernicus.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## R packages

R offers a large variety of methods and packages for visualization. Most plotting methods are however static, which makes sense for publishing results. For model diagnosis, however, interactive plotting can greatly support the modeler. This chapter gives a brief overview of plotting methods that can support you in analyzing your simulation results. 
A great R package to visualize time series data and offer options such as zooming is `dygraphs`. It works with `xts` time series data and can support you in comparing the simulated time series (e.g. discharge, in-stream nitrogen etc.) with observation data. `ggplot2` is a very versatile package for data visualization, that you have seen already several times in the previous chapters. `plotly` offers easy ways to make ggplots more interactive. To visualize spatial data and have functionalities that are usually provided by GIS software `mapview` provides a great alternative. It is based on `leaflet` that most of us have most likely already seen in action on web pages with interactive maps (and maybe did not realize that this is a leaflet map and we can also produce something similar with R easily).

### Package installation

If you do not have installed any of the required R package, follow the instructions for the respective R package. All of the required R packages are available from CRAN and can be installed with the following commands:
```{r, eval=FALSE}
install.packages("dplyr")
install.packages("dygraphs")
install.packages("ggplot2")
install.packages("lubridate")
install.packages("mapview")
install.packages("purrr")
install.packages("plotly")
install.packages("tibble")
install.packages("tidyr")
install.packages("xts")
```

### Loading R packages
```{r, message=FALSE, warning=FALSE}
library(SWATplusR)
library(dplyr)
library(dygraphs)
library(lubridate)
library(tibble)
library(tidyr)
library(purrr)
library(ggplot2)
library(xts)
```

## Loading demo data

The optimization example uses the SWAT+ demo project available from `SWATplusR`. The demo project is a very simple model setups of a head watershed of the Little River Experimental Watershed [LREW; @Bosch2007]. You can load the to your hard drive as follows:

```{r, eval=FALSE}
# The path where the SWAT demo project will be written
demo_path <- "Define:/your/path"

# Loading the SWAT+ demo project on your hard drive
path_plus <- load_demo(dataset = "project",
                       version = "plus",
                       path = demo_path, 
                       revision = 57)
```
```{r, echo=FALSE}
path_plus <- "C:/swat_demo/swatplus_rev57_demo"
```
`SWATplusR` also provides observation data of daily discharge records at the main outlet of the demo for the time period 1968-01-01 until 2012-12-31. We will use the observation data to compare the discharge simulations with observed values:

```{r}
q_obs <- load_demo(dataset = "observation")
```

## Parameter sampling
For the sampling example we will use 7 SWAT model parameters that are frequently calibrated for discharge. Below we create a `tibble` [@Mueller2019] that provides the parameter boundaries for the 7 parameters. We name the parameters using the specific syntax that is requested by the `run_swat*()` functions (see the [Get started](https://chrisschuerz.github.io/SWATplusR/articles/SWATplusR.html#model-parameter-alteration) section on 'Model parameter alteration' to learn more on parameter names):

```{r}
par_bound <- tibble("cn2.hru | change = abschg" = c(-15, 10),
                    "lat_ttime.hru | change = absval" = c(0.5, 50),
                    "lat_len.hru | change = absval" = c(10, 100),
                    "k.sol | change = pctchg" = c(-50, 50),
                    "z.sol | change = pctchg" = c(-50, 50),
                    "esco.hru | change = absval" = c(0, 1),
                    "epco.hru | change = absval" = c(0, 1))
```

### Random sampling with `runif()`
The most basic way to draw random parameter samples is to use the base R function `runif()`. With `runif()` you can create a uniformly distributed vector. To draw a specific number of parameter sets with `runif()` you have to create a vector with that number of random values for each parameter and transform them to their parameter ranges. Below you find a very efficient approach to create a table of parameter combinations using a `map*()` function from the functional programing package `purrr`:

```{r}
n_sample <- 250
par_runif <- map_df(par_bound, ~ runif(n_sample, .x[1], .x[2]))
par_runif
```


### Random sampling with `lhs`

In a very similar way you can draw LHS samples using the function `randomLHS()` from the `lhs` package. Below you find an efficient example to draw a `n_sample = 250` LHS sampled parameter combinations:  

```{r, eval = F}
n_sample <- 250
n_par <- ncol(par_bound)

par_iter1 <- randomLHS(n = n_sample, k = n_par) %>%
  as_tibble(., .name_repair = "minimal") %>%
  map2_df(., par_bound, ~ (.x * (.y[2] - .y[1]) + .y[1])) %>%
  set_names(names(par_bound))
```

```{r, echo=FALSE}
par_iter1 <- readRDS(here::here("vignettes/datasets/par_iter1.rds"))
```

```{r}
par_iter1
```

## Model calibration

### Simulation runs

In a next step we implement all LHS sampled parameter combinations in the SWAT+ demo setup and simulate daily discharges for the calibration period 2003-01-01 until 2007-12-31 (with a warm-up period of 3 years fro 2000 to 2002). To perform the simulations we pass the entire `tibble` with the parameter combinations with the argument `parameter` in the function `run_swatplus()`.

```{r, eval=FALSE}
q_cal1 <- run_swatplus(project_path = path_plus,
                       output = list(q_sim = define_output(file = "channel",
                                     variable = "flo_out",
                                     unit = 1)),
                       parameter = par_iter1,
                       start_date = "2000-01-01",
                        end_date = "2007-12-31",
                       years_skip = 3,
                       n_thread = 4)

#> Building 4 threads in 'Define:/your/path/swatplus_demo/.model_run':
#>  Completed 4 threads in 0S
#> Performing 250 simulations on 4 cores:
#>  Completed 250 simulations in 1M 54S
```

```{r, echo=FALSE}
q_cal1 <- readRDS(here::here("vignettes/datasets/q_iter1.rds"))
```

### Model evaluation

We use the Nash Sutcliffe Efficiency criterion [NSE; @Nash1970] to evaluate the simulations for the calibration period. The NSE is implemented in `hydroGOF` with the function `NSE()`:

```{r}
nse_cal1 <- q_cal1$simulation$q_sim %>%
  select(-date) %>%
  map_dbl(., ~NSE(.x/8.64, q_obs_cal$discharge))
```

The first iteration of discharge simulations and their evaluation with the NSE should help us to identify the relevant parameter and/or constrain parameter boundaries of parameters. Therefore we display the parameter samples with the resulting NSE values in dotty plots:

```{r}
dotty_cal1 <- q_cal1$parameter$values %>%
  mutate(nse = nse_cal1) %>%
  filter(nse > -5) %>%
  gather(key = "par", value = "parameter_range", -nse)

ggplot(data = dotty_cal1) +
  geom_point(aes(x = parameter_range, y = nse)) +
  facet_wrap(.~par, ncol = 3, scales = "free_x") +
  theme_bw()
```

The only parameter that showed a clear parameter front in dotty plot is the "lat_ttime". For the next iteration we will constrain the "lat_ttime" to the interval [0.5,5].

### Parameter update and re-evaluation of the model

To constrain the parameter "lat_ttime" we update the parameter boundaries table. As we only modify the boundaries of one parameter an efficient way is to do the following. In a next step we draw a new set of LHS sampled model parameter combinations and re-run the model:

```{r, eval=FALSE}
par_bound$`lat_ttime.hru | change = absval` <- c(0, 5)

par_cal2 <- randomLHS(n = n_sample, k = n_par) %>%
  as_tibble(.) %>%
  map2_df(., par_bound, ~ (.x * (.y[2] - .y[1]) + .y[1])) %>%
  set_names(names(par_bound))

q_cal2 <- run_swatplus(project_path = path_plus,
                       output = list(q_sim = define_output(file = "channel",
                                     variable = "flo_out",
                                     unit = 1)),
                       parameter = par_iter2,
                       start_date = "2000-01-01",
                        end_date = "2007-12-31",
                       years_skip = 3,
                       n_thread = 4)

#> Building 4 threads in 'Define:/your/path/swatplus_demo/.model_run':
#>  Completed 4 threads in 0S
#> Performing 250 simulations on 4 cores:
#>  Completed 250 simulations in 1M 56S

```

```{r, echo=FALSE}
q_cal2 <- readRDS(here::here("vignettes/datasets/q_iter2.rds"))
```

The updated simulations are again evaluated:
```{r}
nse_cal2 <- q_cal2$simulation$q_sim %>%
  select(-date) %>%
  map_dbl(., ~NSE(.x/8.64, q_obs_cal$discharge))
```

To get an overview of the NSE values for the best simulation runs simply run the following command:

```{r}
sort(nse_cal2, decreasing = T) %>% enframe()
```

In our example we consider all simulations that resulted in an NSE > 0.7 as acceptable simulation results. Therefore we select the acceptable simulations and their respective parameter combinations from our simulation outputs:

```{r}
run_sel <- which(nse_cal2 >= 0.7)
```

In total `r length(run_sel)` simulations resulted in acceptable representations of the observed discharge.

## Model validation

### Parameter sampling and simulation

We will validate the selected `r sum(run_sel)` parameter combinations in the validation period 2008-01-01 to 2012-12-31. Therefore, we use the indices of the acceptable simulations for the calibration period and use their parameter combinations to perform the simulations in the validation period:

```{r, eval=FALSE}
q_val <- run_swatplus(project_path = path_plus,
                      output = list(q_sim = define_output(file = "channel",
                                    variable = "flo_out",
                                    unit = 1)),
                      parameter = q_cal2$parameter,
                      run_index = run_sel,
                      start_date = "2005-01-01",
                       end_date = "2012-12-31",
                      years_skip = 3,
                      n_thread = 4, keep_folder = T)

#> Building 4 threads in 'Define:/your/path/swatplus_demo/.model_run':
#>  Completed 4 threads in 0S
#> Performing 44 simulations on 4 cores:
#>  Completed 44 simulations in 20S
```

```{r, echo=FALSE}
q_val <- readRDS(here::here("vignettes/datasets/q_val.rds"))
```

### Evaluation of the calibration period
The evaluation of the validation period is identical to the evaluation of the iterations of the calibration period. To get an overview of the model performance in the validation period we visualize a summary statistics:

```{r}
nse_val <- q_val$simulation$q_sim %>%
  select(-date) %>%
  map_dbl(., ~NSE(.x/8.64, q_obs_val$discharge))

nse_comp <- tibble(run = names(nse_cal2[run_sel]),
                   calibration = nse_cal2[run_sel],
                   validation = nse_val)

nse_comp %>% 
  gather(key = "period", value = "nse", - run) %>% 
  ggplot(data = .) +
  geom_boxplot(aes(x = period, y = nse), fill = "grey") +
  theme_bw()
```

A tabular overview can show a direct comparison of the simulations in the calibration and validation period for the individual runs and their parametrizations:

```{r}
nse_comp %>% 
  arrange(desc(calibration))
```


## Visualization of discharge simulation

We will plot the ensemble of simulations as a band of minimum and maximum values for each time step and compare it to the observations.

```{r, fig.height= 8}
run_best <- nse_comp$run[1]

get_run_stat <- function(run_table, best_run) {
  run_table %>% 
    select(-date) %>% 
    mutate(q_max = pmap_dbl(., max),
           q_min = pmap_dbl(., min)) %>%
  select(matches(best_run), q_min, q_max) %>%
  set_names(c("q_sim", "q_min", "q_max")) %>%
  add_column(date = run_table$date, .before = 1)
}

cal_stat <- q_cal2$simulation$q_sim %>% 
  select(!!names(q_val$simulation$q_sim)) %>% 
  get_run_stat(., run_best) %>% 
  mutate_if(., is.numeric, ~ (./8.64)) %>% 
  mutate(period = "calibration")

val_stat <- q_val$simulation$q_sim %>% 
  get_run_stat(., run_best) %>% 
  mutate_if(., is.numeric, ~ (./8.64)) %>% 
  mutate(period = "validation")

sim_stat <- bind_rows(cal_stat, val_stat) %>%
  left_join(., q_obs, by = "date") %>%
  rename( q_obs = discharge) %>%
  gather(key = "variable", value = "q", -date, -q_min, - q_max, -period)

ggplot(data = sim_stat) +
  geom_ribbon(aes(x = date, ymin = q_min, ymax = q_max), fill = "grey50") +
  geom_line(aes(x = date, y = q, col = variable), lwd = 0.25) +
  scale_color_manual(values = c("black", "tomato3")) +
  labs( x = "Date (yyyy)", 
        y = expression (Discharge~(m^3~s^{-1})), 
        col = "Discharge") + 
  facet_wrap(period~., scales = "free_x", ncol = 1) + 
  theme_bw() + 
  theme(panel.spacing = unit(0.2, "lines"),
        legend.position = "bottom")
```

## References
