---
title: "Parameter optimization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring SWATplusR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: literature.bib
link-citations: yes
csl: copernicus.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## R packages

Several R packages provide methods for sensitivity analysis. Two of the widely implemented packages are `sensitivity` [@Iooss2018] and `fast` [@Reusser2015]. `fast` provides the Fourier Amplitude Sensitivity Test (FAST) as single method for sensitivity analysis. The implementation of FAST, however, is very well suited to implement it in a workflow with `SWATplusR`. `sensitivity` is a very comprehensive collection of methods for sensitivity analysis (it actually includes the FAST method as well). The implementation in a workflow with `SWATplusR` is, however, less flexible.
In this example we will use two frequently applied methods for sensitivity analysis that are provided by these two R packages, the method of Solbol [-@Sobol1993] provided by `sensitivity` and FAST [@Cukier1973] provided by `fast`.
Most methods for sensitivity analysis require a scalar model result to assess the sensitivity of that output variable to changes in the inputs. Multiple goodness-of-fit functions are available from literature to evaluate simulated time series with observed time series of that variable that result in scalar values. The `hydroGOF` package [@MZB2017] summarizes frequently used functions for the evaluation of time series of hydrological variables.

### Package installation

If you do not have installed any of the required R package, follow the instructions for the respective R package. All of the reuired R packages are available from CRAN and can be installed with the following commands:
```{r, eval=FALSE}
install.packages("sensitivity")
install.packages("fast")
install.packages("hydroGOF")
install.packages("dplyr")
install.packages("tidyr")
install.packages("ggplot2")
```

### Loading R packages
```{r, message=FALSE, warning=FALSE}
library(SWATplusR)
library(sensitivity)
library(fast)
library(hydroGOF)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
```

## Loading demo data

The sensitivity analysis example uses the SWAT+ demo project available from `SWATplusR`. The demo project is a simple model setup of a head watershed of the Little River Experimental Watershed [LREW; @Bosch2007]. You can load the to your hard drive as follows:

```{r, eval=FALSE}
# The path where the SWAT demo project will be written
demo_path <- "Define:/your/path"

# Loading the SWAT+ demo project on your hard drive
path_plus <- load_demo(dataset = "project",
                       swat_version = "plus",
                       path = demo_path)
```

`SWATplusR` also provides observation data of daily discharge records at the main outlet of the demo for the time period 1968-01-01 until 2012-12-31. We will use the observation data to evaluate the model in each optimization step. The model will be evaluated for the time period 2003-01-01 to 2012-12-31. Therefore, we load the demo data and limit it to this time period:

```{r}
q_obs <- load_demo(dataset = "observation")

q_obs <- filter(q_obs, date >= ymd("2003-01-01"),
                       date <= "2012-12-31")
```

## Definition of the function to optimize

Most optimizers require a function as an input argument that uses a parameter set as input and returns a scalar value as a result. To use the `run_swatplus()` function in the optimization we wrap a function around the SWAT model execution that will be implemented in the optimizer to search the optimum parameter set that minimizes the return value (minimization is usually the default option in optimizers). Below you see how you can implement the model execution into a function that can be optimized:

```{r, eval=FALSE}
swat_optim <- function(par, obs) {
  q_sim <- run_swatplus(project_path = "C:/swatplus_demo",
                        output = define_output(file = "channel",
                                              variable = "flo_out",
                                              unit = 1),
                        parameter = par,
                        start_date = "2000-01-01",
                        end_date = "2012-12-31",
                        years_skip = 3,
                        quiet = TRUE)

  nse_q <- - NSE(q_sim$simulation$flo_out/8.64, obs)
  return(nse_q)
}

swat_optim <- function(par, obs) {
  q_sim <- run_swatplus(project_path = "C:/swatplus_demo",
                        output = define_output(file = "channel",
                                              variable = "flo_out",
                                              unit = 1),
                        parameter = par,
                        start_date = "2000-01-01",
                        end_date = "2012-12-31",
                        years_skip = 3,
                        quiet = TRUE)

  nse_q <- - NSE(q_sim$simulation$flo_out/8.64, obs)
  return(nse_q)
  }
```

We defined a function that has one input argument `par` that is the named parameter vector that will be passed to `run_swatplus()`. The simulation will return the discharge at the main outlet (`unit = 1`) for the time period 2003-01-01 to 2012-12-31 (considering the `years_skip = 3` as warm up period). In this case we set the function to be `quiet = TRUE`. In a next step the function uses the observed discharge values from `q_obs$q_out` for the same period and evaluate the simulated discharges using the `NSE()` from the package `hydroGOF` that is then returned as a single value. You can see that we implemented the negative NSE as we will minimize with `optim()`


## References
