---
title: "Exploring SWATplusR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring SWATplusR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: literature.bib
link-citations: yes
linkcolor: blue
csl: copernicus.csl
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Loading R packages

Before exploring the package load `SWATplusR`:
```{r}
library(SWATplusR)
```


## Loading demos

### SWAT projects

`SWATplusR` provides very simple model setups of a head watershed of the Little River Experimental Watershed [LREW; @Bosch2007]. Demo model setups can be retrieved for SWAT2012 and for SWAT+. Currently a SWAT2012 demo is available for Windows and Linux, while for SWAT+ only a setup for Windows is currently available. A demo project can be loaded with the following command.

```{r, eval=FALSE}
# The path where the SWAT demo project will be written
demo_path <- "Define:/your/path"

# Loading a SWAT+ demo project
path_plus <- load_demo(dataset = "project",
                       swat_version = "plus",
                       path = demo_path)

# Loading a SWAT2012 demo project
path_2012 <- load_demo(dataset = "project",
                       swat_version = "2012",
                       path = demo_path)
```

Here `load_demo()` loads a demo project data set and saves it in the defined `demo_path`. Additionally, the function returns the final demo project path (`path_plus` and `path_2012`) as a character string back to R. This path is directly usable to run the SWAT model in that path.

### Observation data
`SWATplusR` also provides observation data of daily discharge records at the main outlet of the demo for the time period 1968-01-01 until 2012-12-31. The data set is useful for the model evaluation of a demo project. You can load the observation data as follows:

```{r}
q_obs <- load_demo(dataset = "observation")

q_obs
```

A quick plot of the observation data set:
```{r}
plot(q_obs, type = "l")
```

### Spatial data
To visualize simulation results spatially, shape files of the subbasins, river network, and HRUs for the SWAT+ and the SWAT2012 demo setups are available from `SWATplusR`. For spatial data `load_demo()` only returns the paths to the shape files. As the spatial data for the two SWAT project setups looks different loading the data requires the definition of the `swat_version`.

```{r}
# Path to the subbasin shape file
sub_path <- load_demo(dataset = "subbasin", swat_version = "plus")
# Path to the subbasin shape file
riv_path <- load_demo(dataset = "river", swat_version = "plus")

sub_path
```

These paths you can use to load the spatial data with your preferred method. The package `sf` [@Pebesma2018a; @Pebesma2018b] provides a very elegant way to work with spatial vector objects in R that have a direct intergration for plotting in `ggplot2` [@Wickham2016].

```{r}
# Loading the sf and ggplot2 packages for the visualization
library(sf)
library(ggplot2)

# Loading the spatial data as simple features
sub <- read_sf(sub_path)
riv <- read_sf(riv_path)

ggplot() + 
  geom_sf(data = sub) + 
  geom_sf(data = riv, col = "royalblue", lwd = 0.75) + 
  theme_bw()

```

## First SWAT model runs

SWAT projects (e.g. one of the demo projects) can be executed from R with the functions `run_swat2012()` to run a SWAT2012 project or `run_swatplus()` to run a SWAT+ project. The minimum requirements to execute a SWAT project are to provide the `project_path` to the project folder on the hard drive and to define the simulation outputs that the simulation should return to R.
The definition of the simulation output always follows a simple pattern. An output variable that should be returned must be defined using the function `define_output()`.
Below is a simple example to run a SWAT+ project and return the `variable = 'flo_out'` that is written into the `file = 'channel'` file for the spatial `unit = 1`. In the file channel the spatial unit relates to the subbasins. **Caution** in the demo SWAT+ model setup the lower subbasin is the subbasin 1.

```{r, echo=FALSE}
path_plus <- "C:/swatplus_demo"
```
```{r, echo=FALSE}
path_2012 <- "C:/swat2012_demo"
```


```{r, eval=FALSE}
q_out <- run_swatplus(project_path = plus_path,
                      output = define_output(file = "channel",
                                             variable = "flo_out",
                                             unit = 1))

#> Building 1 thread in 'Define:/your/path/swatplus_demo/.model_run': 
#>   Thread 1 of 1   Time elapsed: 0S   Time remaining: 0S    
#>   Completed 1 thread in 0S 
#> Performing 1 simulation on 1 core: 
#>   Simulation 1 of 1   Time elapsed: 5S   Time remaining: 0S
#>   Completed 1 simulation in 5S
```

The syntax definition of simulations outputs applies to SWAT+ and SWAT2012. The only difference between the two models in the outdefinition is how to correctly address the output variables. The output files are differenty organized in the two SWAT versions. Below you find an example to extract the simulation outputs of discharge at the main outlet in the SWAT2012 demo setup:

```{r, eval=FALSE}
q_out <- run_swat2012(project_path = path_2012,
                      output = define_output(file = "RCH",
                                             variable = "FLOW_OUT",
                                             unit = 3))

#> Building 1 thread in 'Define:/your/path/swatplus_demo/.model_run': 
#>   Thread 1 of 1   Time elapsed: 3S   Time remaining: 0S    
#>   Completed 1 thread in 3S 
#> Performing 1 simulation on 1 core: 
#>   Simulation 1 of 1   Time elapsed: 4S   Time remaining: 0S
#>   Completed 1 simulation in 4S
```

The discharge in SWAT2012 is written in to the file *"output.rch"*. The discharge leaving a subbasin is written in the column *"FLOW_OUT"*. **Caution**: The unit for the main outlet of the catchment is number 3 in the SWAT2012 setup. The only additional difference in running a basic simuation with SWAT+ and SWAT2012 is the different function call (`run_swatplus()` or `run_swat2012()`).

You might have noticed, that running SWAT also provides information on the progress. This is very convenient for long simulation runs with many model evaluations. If you implement the SWAT function into other function calls the writing to the console can be an undesired behavior. You can quiet the function call with the argument `quiet = TRUE`.

### Output definition




## References
